Sesh: Deadlock-free Sessions with Failure in Rust

### 2019-11-20 doc/ice2019/main.tex
Replaced occurances of Rusty Variation with Sesh

\title{Rusty Variation\\ Deadlock-free Sessions with Failure in Rust}
\author{Wen Kokke
  \institute{LFCS, University of Edinburgh, 10 Crichton St, EH8 9AB, Edinburgh, United Kingdom}
  \email{wen.kokke@ed.ac.uk}
}
\def\titlerunning{Rusty Variation}
\def\authorrunning{Wen Kokke}

@@

\begin{abstract}
  Rusty Variation (RV) is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees. Programs written using RV are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
\end{abstract}

@@

In this paper, we present Rusty Variation\footnote{\url{https://github.com/wenkokke/rusty-variation/}} (RV), a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure. It is these strong guarantees of correctness, together with our account of failure, which sets us apart from previous work~\cite{jespersen2015}. We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}~\cite{rust2019}.

@@

Rusty Variation is an implementation of Exceptional GV (EGV) in Rust. We claim that RV is a faithful implementation of EGV, preserving much of EGV's metatheory, meaning RV guarantees session fidelity, global progress, confluence, and freedom from deadlocks and unwanted races. However, we depart from EGV in two ways:
First, EGV relies upon exceptions and handlers. Rust, however, rejects exceptions in favour of monadic errors, \ie, wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.
Second, the core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee livelock freedom, or---well---termination.

RV is currently implemented on top of the channels provided by the \lstinline{crossbeam_channel} crate, but it can be implemented using any library which offers one-shot channels and selection.

The paper proceeds as follows. In~\cref{sec:contribution}, we describe the programming interface offered by Rusty Variation, and provide several example programs. In~\cref{sec:related-work}, we compare Rusty Variation to the related work by Laumann \etal~\cite{jespersen2015}. Finally, in~\cref{sec:future-work}, we describe future work on verifying the correctness guarantees offered by RV.

\section{Rusty Variation}\label{sec:contribution}

@@

The second program is trickier, as Rust's type system doesn't help us here. Arguably, however, that is for the best. The idea of \emph{linear} session types ignores the reality that sessions may be dropped at any time, for any number of reasons: a process may panic, dropping all sessions in scope, or---when communicating over the network---the connection might fail. In~\cref{sec:introduction}, we mentioned that EGV has an affine type system, and permits the explicit cancellation of sessions. Following EGV, Rusty Variation also permits the cancellation of sessions. However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/1.35.0/book/ch15-03-drop.html}}.

@@

RV offers an alternative, the \lstinline{select_mut} function. This function operates on a vector of \emph{receiving} endpoints, returning the result of the first receive operation which completes, and removing the corresponding session from the vector. We restrict ourselves to the receiving endpoints, as RV's \lstinline{send} operation is non-blocking, hence selecting on sending endpoints doesn't make any sense.
\begin{lstlisting}
pub fn select_mut<T, S: Session>(rs: &mut Vec<Recv<T, S>>) -> Option<(T, S)>
\end{lstlisting}

The \lstinline{select_mut} function is implemented on top of \lstinline{crossbeam_channel::Select}, in a way which respects the rules listed above. In addition, RV offers an immutable variant of \lstinline{select_mut}, named \lstinline{select}, which returns the remainder of the endpoints along the with result of the selection. Using \lstinline{select_mut}, we can write the following program, which forks off ten threads, each sending an index between 1 and 10, and then repeatedly selects on the receiving endpoints, receiving and printing an index, printing the numbers 1 to 10 in a random order:

@@

\subsection{Implementation}\label{sec:implementation}
So far we have avoided discussing the implementation whenever possible, and with good reason: RV can be implemented using \emph{any} library which offers one-shot channels and selection. Hence, the important contribution in this paper is not the implementation, but the programming interface offered by RV.

@@

The Rust standard library offers \lstinline{std::sync::mpsc}, an implementation of channels. The first implementation of RV used the channels from this library. However, \lstinline{std::sync::mpsc} has some fundamental issues with its design\footnote{\url{https://github.com/rust-lang/rust/pull/42397\#issuecomment-315867774}}, and seems to be on its way out. In 2018, the selection mechanism, \lstinline{mpsc_select!}, was deprecated\footnote{\url{https://github.com/rust-lang/rust/issues/27800}}, and there are plans to phase out \lstinline{std::sync::mpsc} altogether~\cite{glavina2019}.

\subsection{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Rusty Variation lock in safe Rust. The first is by writing an infinite loop, which we discussed earlier.

@@

The following two programs illustrate the difference in handling failure between Rusty Variation and the work by Laumann \etal~\cite{jespersen2015}. In both programs, the programmer has forgotten to complete the server, accidentally dropping the session endpoint \lstinline{_s} or channel \lstinline{_c}, respectively. The program on the left, written in RV, returns \lstinline{None}. The program on the right, written using the library described by Laumann \etal\cite{jespersen2015}, segfaults. We have since reported this segfault to Laumann \etal, and they have fixed it in the most recent version of the library. The program on the right now panics:

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Rusty Variation
// main() returns None
fn server(_s: Recv<(), End>) {
  // unintentionally left blank.
}
fn client(s: Send<(), End>) {
  close(send((), s))
}
fn main() -> Option<()> {
  client(fork(server))
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
// main() causes segfault
fn server(_c: Chan<(), Recv<(), Eps>>) {
  // unintentionally left blank.
}
fn client(c: Chan<(), Send<(), Eps>>) {
  c.send(()).close();
}
fn main() -> () {
  connect(server, client);
}
\end{lstlisting}
\end{minipage}
\vspace{-0.25\baselineskip}

While we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices. Below we show two definitions of the session type for a stream of integers, one in Rusty Variation, and one using the library by Laumann \etal~\cite{jespersen2015}:

\begin{minipage}{0.5\linewidth}
\begin{lstlisting}
// Rusty Variation
type IntStream = Send<i32, IntStream>;
\end{lstlisting}  
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
type IntStream = Rec<Send<i32, Var<Z>>>;
\end{lstlisting}  
\end{minipage}
\vspace{-0.25\baselineskip}

@@

A last feature which distinguishes Rusty Variation from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice. Rusty Variation uses enumerations to provide labelled choice, as described in~\cref{sec:choice}. Laumann \etal~\cite{jespersen2015} provide an \lstinline{offer!} macro, which expands a labelled offer to a series of binary offers, \eg offering a ternary choice expands to two nested invocations of \lstinline{offer_either}.

@@

\subsection{Testing \& Verification}\label{sec:verification}
In~\cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.

Firstly, RV is unit tested. We test various use-cases from EGV, from basic operations and their interactions with cancellation, to delegation and recursive session types. This test suite has caught several errors in the implementation. However, unit tests are error prone, and it is hard (if not impossible) to obtain a set of tests which cover all use-cases of a library. Unfortunately, tools such as RustBelt~\cite{jung2017} and Rust Distilled~\cite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of RV.

In the absence of a mature formal semantics for Rust, we are designing a formal language, RV (or ``formal RV''), which matches, as close as possible, the semantics of our Rust library. It is a concurrent lambda calculus with heap-based shared-memory semantics. We plan to prove that formal RV guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom. Furthermore, we plan to show that there exists a translation from EGV to formal RV, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\cite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV. This would show that formal RV is a faithful implementation of EGV.

Lastly, we plan to ensure the correspondence between formal RV and the implementation of RV, using property-based testing. We plan to generate a series of random formal RV terms, using Neat~\cite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal RV.

\subsection{Extending Exceptional GV}
Several features of Rusty Variation are not covered by the formal language EGV. In~\cref{sec:selection}, we mention that selection isn't a part of EGV. In~\cref{sec:related-work}, we mention that recursion isn't part of EGV. If we wish to have any hope of formally verifying RV, along the lines described in~\cref{sec:verification}, we would need to extend EGV to cover these features.

>>>>

\title{Sesh: Deadlock-free Sessions with Failure in Rust}
\author{Wen Kokke
  \institute{LFCS, University of Edinburgh, 10 Crichton St, EH8 9AB, Edinburgh, United Kingdom}
  \email{wen.kokke@ed.ac.uk}
}
\def\titlerunning{Sesh}
\def\authorrunning{Wen Kokke}

@@

\begin{abstract}
  Sesh is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees. Programs written using Sesh are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
\end{abstract}

@@

In this paper, we present Sesh\footnote{\url{https://github.com/wenkokke/sesh/}}, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure. It is these strong guarantees of correctness, together with our account of failure, which sets us apart from previous work~\cite{jespersen2015}. We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}~\cite{rust2019}.

@@

Sesh is an implementation of Exceptional GV (EGV) in Rust. We claim that Sesh is a faithful implementation of EGV, preserving much of EGV's metatheory, meaning Sesh guarantees session fidelity, global progress, confluence, and freedom from deadlocks and unwanted races. However, we depart from EGV in two ways:
First, EGV relies upon exceptions and handlers. Rust, however, rejects exceptions in favour of monadic errors, \ie, wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.
Second, the core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee livelock freedom, or---well---termination.

Sesh is currently implemented on top of the channels provided by the \lstinline{crossbeam_channel} crate, but it can be implemented using any library which offers one-shot channels and selection.

The paper proceeds as follows. In~\cref{sec:contribution}, we describe the programming interface offered by Sesh, and provide several example programs. In~\cref{sec:related-work}, we compare Sesh to the related work by Laumann \etal~\cite{jespersen2015}. Finally, in~\cref{sec:future-work}, we describe future work on verifying the correctness guarantees offered by Sesh.

\section{Sesh}\label{sec:contribution}

@@

The second program is trickier, as Rust's type system doesn't help us here. Arguably, however, that is for the best. The idea of \emph{linear} session types ignores the reality that sessions may be dropped at any time, for any number of reasons: a process may panic, dropping all sessions in scope, or---when communicating over the network---the connection might fail. In~\cref{sec:introduction}, we mentioned that EGV has an affine type system, and permits the explicit cancellation of sessions. Following EGV, Sesh also permits the cancellation of sessions. However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/1.35.0/book/ch15-03-drop.html}}.

@@

Sesh offers an alternative, the \lstinline{select_mut} function. This function operates on a vector of \emph{receiving} endpoints, returning the result of the first receive operation which completes, and removing the corresponding session from the vector. We restrict ourselves to the receiving endpoints, as Sesh's \lstinline{send} operation is non-blocking, hence selecting on sending endpoints doesn't make any sense.
\begin{lstlisting}
pub fn select_mut<T, S: Session>(rs: &mut Vec<Recv<T, S>>) -> Option<(T, S)>
\end{lstlisting}

The \lstinline{select_mut} function is implemented on top of \lstinline{crossbeam_channel::Select}, in a way which respects the rules listed above. In addition, Sesh offers an immutable variant of \lstinline{select_mut}, named \lstinline{select}, which returns the remainder of the endpoints along the with result of the selection. Using \lstinline{select_mut}, we can write the following program, which forks off ten threads, each sending an index between 1 and 10, and then repeatedly selects on the receiving endpoints, receiving and printing an index, printing the numbers 1 to 10 in a random order:

@@

\subsection{Implementation}\label{sec:implementation}
So far we have avoided discussing the implementation whenever possible, and with good reason: Sesh can be implemented using \emph{any} library which offers one-shot channels and selection. Hence, the important contribution in this paper is not the implementation, but the programming interface offered by Sesh.

@@

The Rust standard library offers \lstinline{std::sync::mpsc}, an implementation of channels. The first implementation of Sesh used the channels from this library. However, \lstinline{std::sync::mpsc} has some fundamental issues with its design\footnote{\url{https://github.com/rust-lang/rust/pull/42397\#issuecomment-315867774}}, and seems to be on its way out. In 2018, the selection mechanism, \lstinline{mpsc_select!}, was deprecated\footnote{\url{https://github.com/rust-lang/rust/issues/27800}}, and there are plans to phase out \lstinline{std::sync::mpsc} altogether~\cite{glavina2019}.

\subsection{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Sesh lock in safe Rust. The first is by writing an infinite loop, which we discussed earlier.

@@

The following two programs illustrate the difference in handling failure between Sesh and the work by Laumann \etal~\cite{jespersen2015}. In both programs, the programmer has forgotten to complete the server, accidentally dropping the session endpoint \lstinline{_s} or channel \lstinline{_c}, respectively. The program on the left, written in Sesh, returns \lstinline{None}. The program on the right, written using the library described by Laumann \etal\cite{jespersen2015}, segfaults. We have since reported this segfault to Laumann \etal, and they have fixed it in the most recent version of the library. The program on the right now panics:

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Sesh
// main() returns None
fn server(_s: Recv<(), End>) {
  // unintentionally left blank.
}
fn client(s: Send<(), End>) {
  close(send((), s))
}
fn main() -> Option<()> {
  client(fork(server))
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
// main() causes segfault
fn server(_c: Chan<(), Recv<(), Eps>>) {
  // unintentionally left blank.
}
fn client(c: Chan<(), Send<(), Eps>>) {
  c.send(()).close();
}
fn main() -> () {
  connect(server, client);
}
\end{lstlisting}
\end{minipage}
\vspace{-0.25\baselineskip}

While we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices. Below we show two definitions of the session type for a stream of integers, one in Sesh, and one using the library by Laumann \etal~\cite{jespersen2015}:

\begin{minipage}{0.5\linewidth}
\begin{lstlisting}
// Sesh
type IntStream = Send<i32, IntStream>;
\end{lstlisting}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
type IntStream = Rec<Send<i32, Var<Z>>>;
\end{lstlisting}
\end{minipage}
\vspace{-0.25\baselineskip}

@@

A last feature which distinguishes Sesh from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice. Sesh uses enumerations to provide labelled choice, as described in~\cref{sec:choice}. Laumann \etal~\cite{jespersen2015} provide an \lstinline{offer!} macro, which expands a labelled offer to a series of binary offers, \eg offering a ternary choice expands to two nested invocations of \lstinline{offer_either}.

@@

\subsection{Testing \& Verification}\label{sec:verification}
In~\cref{sec:contribution}, we claim that Sesh preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of Sesh is out of the scope of this paper, it would be remiss of us not substantiate this claim.

Firstly, Sesh is unit tested. We test various use-cases from EGV, from basic operations and their interactions with cancellation, to delegation and recursive session types. This test suite has caught several errors in the implementation. However, unit tests are error prone, and it is hard (if not impossible) to obtain a set of tests which cover all use-cases of a library. Unfortunately, tools such as RustBelt~\cite{jung2017} and Rust Distilled~\cite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of Sesh.

In the absence of a mature formal semantics for Rust, we are designing a formal language, Sesh (or ``formal Sesh''), which matches, as close as possible, the semantics of our Rust library. It is a concurrent lambda calculus with heap-based shared-memory semantics. We plan to prove that formal Sesh guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom. Furthermore, we plan to show that there exists a translation from EGV to formal Sesh, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\cite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV. This would show that formal Sesh is a faithful implementation of EGV.

Lastly, we plan to ensure the correspondence between formal Sesh and the implementation of Sesh, using property-based testing. We plan to generate a series of random formal Sesh terms, using Neat~\cite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal Sesh.

\subsection{Extending Exceptional GV}
Several features of Sesh are not covered by the formal language EGV. In~\cref{sec:selection}, we mention that selection isn't a part of EGV. In~\cref{sec:related-work}, we mention that recursion isn't part of EGV. If we wish to have any hope of formally verifying Sesh, along the lines described in~\cref{sec:verification}, we would need to extend EGV to cover these features.
